### 2020-05-15 ThesisJo/2_background.tex
2020-04-26
Auto-commit

@@

% Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\
\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Method relations. This thesis explores three beamformers reconstructing images from acoustic echoes: element-space MVDR (reference), beam-space MVDR (close approximation) and element-space LCA (educated trial-and-error). Article IV presents background on synthesis of image templates from reconstructed images. }%
\label{2_fig_article_relation}
\end{figure}

@@

If $f$ represents displacement of a vibrating string, (\ref{eq_wave_equation}) states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side). In acoustics $f$ represents sound pressure, with SI unit Pascal [Pa = N/m$^2$] or logarithmic unit sound-pressure-level (SPL) [1\,Pa = 94\,$\mathrm{dB}_{\textrm{SPL}}$]. To generalize and simplify, we keep our equations free of physical units and simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.

The linear homogeneous wave equation has a few solutions of interest, in particular that of a plane and spherical wave. For a field emitted from source position $\udot{S}$ and observed from $m$th sensor position $\udot{m}$, the solutions are
%
\begin{align}
f_\textrm{plane}(t,\vec{p}_{mS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{mS})} \\
f_\textrm{spheric}(t,\vec{p}_{mS}) &= \frac{A}{\lVert \vec{p}_{mS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{mS} \rVert)},
\end{align}
% 
where $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{mS}$ is position vector $\udot{m}$ to $\udot{S}$. We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to wavelength and array length) and as superimposed spherical waves under near-field conditions.

% One is that of plane wave and the other a  A signal source with the position $\udot{S}$ . We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,
% 
% An introduction to such notation can be found in \chapref{IV_methods}{IV}.
% 
% 
% reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length)

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (Fig. \ref{2_fig_article_relation}). Further assume the sensors are linear, infinite bandwidth and invarant to shifts in position (homogeneous, inter-independent), direction (isotropic) and time. Image reconstruction coherently combines these receiver outputs to focus at an angle or point at a time. Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$. The delays (normalized by sampling frequency $f_s$) coarsely aligns signals from source $\udot{S}$,
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c\,f_s},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$. Subtracting the minimum distance is mathematically optional, but facilitates physical realization and low latency. By assuming implicit dependence on $\vec{p}$ and data pre-delayed to the pixel of interest, we simplify notation to $x_m[n]\in\mathbb{C}$. Beamformer output $z[n]\in\mathbb{C}$ is, by definition, the weighted sum of delayed data samples,
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where the weights are normalized and $w_m=\in\mathbb{C}$ is the weight factor assigned to channel $m$. The weights are normalized, $\dvec{w}^H\dvec{1}=1$, Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_{Am}},\label{2_eq_beam pattern}
\end{align}
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
where the start position $\udot{A}$ is arbitrary. Note how the beam pattern depends on both angle and frequency of the incoming wave.

The beamformer is called delay-and-sum (DAS) or conventional if the weights are static (temporally non-varying). Various weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. A class of adaptive beamformers target this limitation by adapting the weights to the dynamic impinging wavefield. One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}. We discuss two versions of it termed by

@@

\subsection{Element-space MVDR}

@@

Element-space (time-domain) MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile (more on this later). It computes the weights that minimize noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Solving this convex optimization problem with Lagrange multipliers yields
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{a}=\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. In the presence of zero-mean random Gaussian noise, these weights yield the maximum likelihood estimate of the signal and the maximum signal-to-interference-and-noise ratio (SINR). However, active systems or passive ones subjected to multipath or jamming, experience highly correlated noise, interference and signal~\cite{Widrow1982}. Limited amount of data, array miscalibration and model inaccuracies makes matters worse, sometimes resulting in performance inferior to DAS~\cite{Li2006robust}.

To make MVDR robust, it must avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness. Literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
%
\begin{itemize}
\item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays. We achieve this through subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure. Some studies construct inversion-friendly Toeplitz-matrices directly [TK87]. $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89]. 
\item \emph{Temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
\item \emph{Diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%
%
These techniques ensure a numerically well conditioned $\eR$, allow reducing the computational complexity and expose enough data parallelism to facilitate GPU-acceleration.

Topeplitz [TK87] - choose wm to make the smoothed matrix close to Toeplitz. Topelitz weighting, adaptive spatial averaging in literature. [TK87]

@@

\subsection{Beam-space MVDR}

Systems with narrow beams, narrow bands, or both, have most backscattered energy residing in center spatial frequencies. Zero-frequency maps to broadside, higher frequencies map to larger opening angles. Computing weighted combinations of such frequencies, similarly to doing so with elements, is referred to as beamspace or discrete Fourier-transform (DFT) beamforming.

@@

In imaging sonar and ultrasound systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}. Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation. We mitigate this by applying space reduction in the spatial frequency-domain, or beamspace.

Beam-space MVDR resembles element-space MVDR, but demands an initial Fourier transformation of the data:
%
\begin{gather}
\x_{\text{BS},l}[n] = \mat{B}\x_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L},\label{2_eq_beamspace}
\end{gather}
%
where $l$ is subarray index, $N_b$ is number of beams that should be computed and
%
\begin{gather}
[\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{2_eq_butler}
\end{gather}
%
is the Butler (or DFT) matrix~\cite{Butler1961}. Once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\w_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).

@@

\subsection{LCA beamformer}

@@

\section{Image synthesis}

@@

>>>>

This chapter

@@

% Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\
\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Method relations. This thesis explores three beamformers reconstructing images from acoustic echoes: element-space MVDR (reference), beam-space MVDR (close approximation) and element-space LCA (educated trial-and-error). Article IV presents background on synthesis of image templates from reconstructed images. }\label{2_fig_article_relation}
\end{figure}

@@

If $f$ represents displacement of a vibrating string, (\ref{eq_wave_equation}) states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side). In acoustics $f$ represents sound pressure, with SI unit Pascal [Pa = N/m$^2$] or logarithmic unit sound-pressure-level (SPL) [1\,Pa = 94\,$\mathrm{dB}_{\textrm{SPL}}$]. To generalize and simplify, we free our equations of physical units and simply think of $f$ as a signal: a unit-less scalar function with wave characteristics. Consequently, every presented image will be unit-less as well.

The linear homogeneous wave equation has a few solutions of interest, in particular that of a plane and spherical wave. For a field emitted from source position $\udot{S}$, observed from the $m$th sensor at position $\udot{m}$, the solutions are
%
\begin{align}
f_\textrm{plane}(t,\vec{p}_{mS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{mS})} \\
f_\textrm{spheric}(t,\vec{p}_{mS}) &= \frac{A}{\lVert \vec{p}_{mS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{mS} \rVert)},
\end{align}
% 
where $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{mS}$ is position vector $\udot{m}$ to $\udot{S}$. We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to wavelength and array length) and as superimposed spherical waves under near-field conditions.

% One is that of plane wave and the other a  A signal source with the position $\udot{S}$ . We can represent the field observessd at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,
% 
% An introduction to such notation can be found in \chapref{IV_methods}{IV}.
% 
% 
% reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length)

@@

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (\Fig{2_fig_article_relation}). Further assume the sensors are linear, infinite bandwidth and invarant to shifts in position (homogeneous, inter-independent), direction (isotropic) and time. Image reconstruction coherently combines these receiver outputs to focus at an angle or point at a time. Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (\Fig{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\in\mathbb{V}^3$ its absolute orientation. Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$. The delays (normalized by sampling frequency $f_s$) coarsely aligns signals from source $\udot{S}$,
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c\,f_s},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$. Subtracting the minimum distance is mathematically optional, but facilitates physical realization and low latency. Phase shifts can substitute delays in applications that are narrow band, narrow beam, or both~\cite{Johnson1993}. By assuming implicit dependence on $\vec{p}$ and data pre-delayed to the pixel of interest, we simplify notation to $x_m[n]\in\mathbb{C}$. Beamformer output $z[n]\in\mathbb{C}$ is, by definition, the weighted sum of delayed data samples,
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where the weights are normalized and $w_m=\in\mathbb{C}$ is the weight factor assigned to channel $m$. %The weights are normalized, $\dvec{w}^H\dvec{1}=1$, \todo{???} 
Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_{Am}},\label{2_eq_beam pattern}
\end{align}
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
where the start position $\udot{A}$ is arbitrary. Note how the beam pattern depends on both angle and frequency of the incoming wave.

The beamformer is called delay-and-sum (DAS) or conventional if the weights are static (temporally non-varying). Various weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. Adaptive beamformers target this limitation by adapting the weights, to spatial and temporal features of the dynamically impinging wavefield.

The minimum variance distortionless response (MVDR) or Capon beamformer (named after its inventor for use in spectral estimation~\cite{Capon1969}), analytically yields array weights from a continuous solution space that are optimal in the minimum variance sense. Section \ref{sec_es_mvdr} describes its formulation in the temporal domain, used in

 %We discuss two versions of it termed by \todo{???}

@@

\subsection{Element-space MVDR}\label{sec_es_mvdr}

@@

Element-space (time-domain) MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile (more on this later). It computes the weights that minimize noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. We refer to it as fully adaptive because it  Solving this convex optimization problem with Lagrange multipliers yields
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{a}=\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. In the presence of zero-mean random Gaussian noise, these weights yield the maximum likelihood estimate of the signal and the maximum signal-to-interference-and-noise ratio (SINR). However, active systems or passive ones subjected to multipath or jamming, experience highly correlated noise, interference and signal~\cite{Widrow1982}. Limited amount of data, array miscalibration and model inaccuracies makes matters worse, sometimes resulting in performance inferior to DAS~\cite{Li2006robust}.

To make MVDR robust, it must avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness. The literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
%
\begin{itemize}
\item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays. We achieve this through subarray averaging~\cite{Kailath1985,Johnson1993}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure. Some studies construct inversion-friendly Toeplitz-matrices directly [TK87]. $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89]. 
\item \emph{Temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
\item \emph{Diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%
%
These techniques ensure a numerically well conditioned $\eR$, allow reducing the computational complexity and expose enough data parallelism to facilitate GPU-acceleration.

\todo{???}
Topeplitz [TK87] - choose wm to make the smoothed matrix close to Toeplitz. Topelitz weighting, adaptive spatial averaging in literature. [TK87]

@@

\subsection{Beam-space MVDR}\label{sec_bs_mvdr}

Systems with narrow beams, narrow bands, or both, have most backscattered energy residing in center spatial frequencies. Zero-frequency maps to broadside, higher frequencies map to larger opening angles. Beamspace (discrete Fourier-transform, DFT) beamforming coherently combines such frequencies in a weighted manner, similarly to doing so with elements.

@@

In sonar and ultrasound imaging systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}. Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation. We mitigate this by applying space reduction in the spatial frequency-domain, or beamspace.

Beam-space MVDR resembles element-space MVDR, but demands an initial Fourier transformation of the data:
%
\begin{gather}
\dvec{x}_{\text{BS},l}[n] = \mat{B}\dvec{x}_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L},\label{2_eq_beamspace}
\end{gather}
%
where $l$ is subarray index, $N_b$ is number of beams that should be computed and
%
\begin{gather}
[\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}\label{2_eq_butler}
\end{gather}
%
is the Butler (or DFT) matrix~\cite{Butler1961}. Once $\dvec{x}_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\dvec{w}_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).

@@

\subsection{LCA beamformer}\label{sec_lca}

@@

\section{Image synthesis}\label{sec_image_synthesis}

@@

\quoteload{IV_text_introduction}

### 2020-05-15 ThesisJo/3_publications.tex
Auto-commit

We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three \emph{reconstruct} images from echoes and the last \emph{synthesizes} through simulation.

\begin{figure}[tp]\label{III_fig_article_relation}
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Article relations (repost of Figure \ref{I_fig_article_relation} in chapter 1). The first three articles reconstructs images from acoustic echoes. The last article synthesizes image templates from 3D models and object parameters estimated from the reconstructed image. Accelerated adaptivity is key: it enables rapid dynamic responses to mission events. }
\end{figure}

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\section{Paper I}\label{sec:paperI} %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\textbf{An Optimized GPU implementation of the\\MVDR Beamformer for Active Sonar Imaging}~\cite{Buskenes2014}\\
J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng\\
\textit{IEEE Jou\textbf{}rnal of Oceanic Engineering}\\
Volume 40, Issue 2, April 2015

\textit{In short}:
\begin{enumerate}
\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
\end{enumerate}

This article accelerates element-space MVDR with GPUs for use in active sonar imaging. Element-space MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity. We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense. Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.

@@

We find that implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.

@@

@@

@@

@@

\textit{In short}:
\begin{enumerate}
\item Implements a SAS simulator on GPU that allows AUVs to compare its images on-the-fly to vast numbers of objects and configurations, yielding superior classification results and real-time responses.
\item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar %, and tied with it above. 
\end{enumerate}

>>>>

We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic imagery: the first three \emph{reconstruct} images from echoes and the last \emph{synthesizes} images through simulation.

This chapter briefly describes the motivation, methods and results of each article. \Fig{3_fig_article_relation} (same as \Fig{2_fig_article_relation}) offers visual context.

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Article relations (repost of Figure \ref{2_fig_article_relation}). The first three articles reconstructs images from acoustic echoes. The last article synthesizes image templates from 3D models and object parameters estimated from the reconstructed image. Accelerated adaptivity enables rapid dynamic responses to mission events. }\label{3_fig_article_relation}
\end{figure}

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\newpage
\section{Paper I}\label{sec:paperI} %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% Motivation
% Methods
% Results

\textbf{An Optimized GPU implementation of the\\MVDR Beamformer for Active Sonar Imaging}~\cite{Buskenes2014}\\
J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng\\
\textit{IEEE Jou\textbf{}rnal of Oceanic Engineering}\\
Volume 40, Issue 2, April 2015

\textit{Key results}
\begin{enumerate}
\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
\item Offers rarely documented insights into GPU hardware limitations, design implications and computational efficiency.
\end{enumerate}

%Element-space MVDR beamforming on GPUs in active sonar imaging

Element-space MVDR is often considered a reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity. Unfortunately, its computational complexity is high because a spatial covariance matrix must be estimated and inverted for each image pixel. This discourages its use in sonar systems aiming at longer ranges, higher resolutions and real-time performance---all competing for the same computational resources.

@@

Implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.

@@

In this study we show that for active sonar systems up to 32 channels, the computation time
can be significantly reduced by performing arithmetic optimizations, and by implementing
the MVDR beamformer on a graphics processing unit (GPU). We point out important
hardware limitations for these devices, and assess the design in terms of how efficiently it is
able to use the GPU’s resources. On a quad-core Intel Xeon system with a high-end Nvidia
GPU, our GPU implementation renders more than a million pixels per second (1 MP/s).
Compared to our initial central processing unit (CPU) implementation the optimizations
described herein led to a speedup of more than two orders of magnitude, or an expected
five to ten times improvement had the CPU received similar optimization effort. This
throughput enables real-time processing of sonar data, and makes the MVDR a viable
alternative to conventional methods in practical systems.

@@

Beam-space MVDR beamforming on GPUs in clinical ultrasound imaging

@@

LCA beamforming versus MVDR in active sonar imaging

@@

Real-time sonar image synthesis on GPUs for automatic target recognition

\textit{In short}:
\begin{enumerate}
\item Implements a SAS simulator on GPU that allows AUVs to compare its images on-the-fly to vast numbers of objects and configurations, yielding superior classification results and real-time responses.
\item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar\todo{???} %, and tied with it above. 
\end{enumerate}

### 2020-05-15 ThesisJo/4_discussion.tex
Auto-commit

%&header
% !TeX root = 3_discussion
\input{subheader}
\endofdump

@@

\chapter{Discussion and Outlook}

The main contributions of this thesis, along with brief discussion and onlook follows.

\section{Summary}

\textbf{Paper I} - \textit{Element-space MVDR beamforming on GPUs in active sonar imaging}:
\begin{enumerate}
\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
\end{enumerate}

\textbf{Paper II} - \textit{Beam-space MVDR beamforming on GPUs in clinical ultrasound imaging}:
\begin{enumerate}
\item Alters the implementation in Paper I to operate in beam-space (frequency domain), reduced to its three center beams (the most significant ones).
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with 64 and 96 channels.
\item Yields similar image quality to element-space MVDR, in accordance with former studies in active sonar.
\end{enumerate}

\textbf{Paper III} - \textit{LCA beamforming versus MVDR in active sonar imaging}:
\begin{enumerate}
\item Shows that LCA beamformer performs similarly to MVDR with only 6 weight sets to choose from: The rectangular and Kaiser function with $\beta=6$, steered to $\phi\in[0,\pm\frac{1}{2}\delta{\phi_{-3dB}}]$, where $\delta{\phi_{-3dB}}$ is the $-3$\,dB bandwidth of the amplitude response to the weight set.
%Contrasts the LCA and MVDR beamformer, two different ways to solve the minimum variance criterion. The former selects the best performing weights from a predefined set, the latter computes them from a sample covariance matrix estimated from the input data.
\item Finds no further improvement above 15 suitable weight sets.
\item Provides insights into typical amplitude responses of MVDR constrained for active sonar imaging and pinpoints their resemblance to the chosen LCA weight sets. 
\item Suggests swapping the Kaiser function trigonometric one parametrized by $\alpha$, yielding similar image quality while allowing an analytic solution for $\alpha$ to be found. This reduces the algorithm to simply iterating through steering angles.
\end{enumerate}

\textbf{Paper IV} - \textit{Real-time sonar image synthesis on GPUs for automatic target recognition}:
\begin{enumerate}
\item Implements a SAS simulator on GPU that allows AUVs to compare its images on-the-fly to vast numbers of objects and configurations, yielding superior classification results and real-time responses.
\item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar %, and tied with it above. 
\end{enumerate}

\section{Discussion}

MVDR yield optimal weights in the minimum variance sense but is computationally complex and user unfriendly. Parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.

@@

@@

\section{Onlook}

Although receive adaptive beamforming yields slight image quality improvement in our active imaging systems, we believe its transmit potential to be even greater. Ideally the two should work in synergy, receiver and transceiver cooperating in pursuit of common goals, for example to suppress dynamic interference. Low latency feedback loops is key to making this work.

Software defined radio 
Applications 5G, Wifi,

\emph{Phased arrays}

\emph{System on Chips} (SoCs) are heterogeneous computing platforms crafted from a single silicon chip. One of these chips, the Ultrascale+ Radio Frequency SoC (RFSoC) from Xilinx, packs a quad-core CPU, dual-core real-time processor, GPU, Field Programmable Gate Array (FPGA) and up to 16 Analog-to-Digital and Digital-to-Analog Converters (ADCs and DACs)---all in one silicon chip. It facilitates multiple-input-multiple-output (MIMO) and other phased-array  can easily replace most analog front-end electronics in a sonar or ultrasound system with a software-defined solution, yielding better flexibility, reduced design complexity, improved bandwidth and lowered latency, all in a tiny physical footprint.

>>>>

%&header
% !TeX root = thesis
% !TeX root = 3_discussion
\input{subheader}
\endofdump

@@

\chapter{Discussion}

The main contributions of this thesis, along with brief discussion follows.

MVDR yields optimal weights in the minimum variance sense but is computationally complex and user unfriendly. Parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.

@@

Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.

@@
